Analysing Constraint Grammars with a SAT-solver

### 2016-03-03 paper/lrec/problem_desc.tex

Constraint Grammar (CG, \newcite{karlsson1995constraint})
% \citep[CG,][]{karlsson1995constraint} 
is a formalism used to disambiguate morphologically analysed text. 
A grammar consists of rules that target specific readings for selection or removal, based on contextual tests. For example, the following rule
\begin{itemize}
\item[] \texttt{REMOVE verb IF (-1 det) ;}
\end{itemize}
removes all verb readings from a word which is preceded by a determiner.
Given the following text,
\begin{itemize}
\item[] 
\begin{verbatim}
"<the>"
        "the" det def
"<wish>"
        "wish" noun sg
        "wish" verb pl
        "wish" verb inf
\end{verbatim}
\end{itemize}
the rule will match to the word \emph{wish}, and remove the readings
\texttt{"wish" verb pl} and \texttt{"wish" verb inf}.
However, if the target word has only one remaining reading, then the
rule will not apply, even if the condition is met.

CGs are valuable resources for rule-based NLP, especially for lesser
resourced languages. They are robust and can be written without large
corpora---only morphological analyser is needed. The formalism is
lightweight and language-independent, and resources can be shared
between related languages \cite{bick2006spanish,lene_trond_linda2010}.
Mature CGs contain some thousands of rules, but even small CGs are
shown to be effective \cite{lene_trond2011}.

@@

\begin{itemize}
\item Are all the rules distinct? (e.g. \texttt{Para} and \texttt{De} may be included in \texttt{Prep})
\item Can two or more rules be merged? (e.g. \texttt{SELECT Inf IF
    (-1C Prep OR Vai OR Vbmod)})
\item Can I rewrite a messy rule in a neater way without changing the meaning?
\item What is the best order for the rules?
\item Generate a sequence that triggers $R$ but not $R'$ 
\end{itemize}

@@

For the second use case, here are examples of conflicts that our tool will detect.
\begin{itemize}
\item If a rule appears twice, the second occurrence will be disabled by the first
\item $r$ selects something in a context, $r'$ removes~it
\item $R$ removes something from the context of $r$, so $r$ can never
  apply
\item $r$ has an internal conflict, such as non-existent
tag combination, or contradictory requirements for a context word
\end{itemize}
$R$ can be a single rule or a set of rules: for instance, if one rule removes a verb in
context $C$, and another in context $\neg C$, together these rules
remove a verb in all possible cases, disabling any future rule that
targets verbs.

@@

We express these constraints as a \emph{Boolean satisfiability problem} (SAT).
A SAT-problem consists of two components: a set of Boolean variables, and a set
of clauses on those variables. For instance, let the set ov variables
be $\{a, b\}$ and the formulas $\{a \vee b, \neg{}a\}$. A program called
\emph{SAT-solver} will try to find a solution, where all the variables
are replaced by a Boolean value. For this particular problem, the
unique solution is $\{a=False, b=True\}$, but it is also possible for a
SAT-problem to have no solution or multiple solutions.
In the case of CG analysis, the model we build is a \emph{sentence},
which starts as potentially anything, and is being shaped by all the
rules along the way. This is how we can generate possible inputs, as
well as check if the grammar is internally consistent.

The paper is structured as follows. Section~\ref{sec:prev} relates our
work to the previous research. 
Section~\ref{sec:implementation} discusses the implementation, and
Section~\ref{sec:eval} presents preliminary results. 
Section~\ref{sec:conclusion} discusses future work and concludes the paper.

>>>>

Constraint Grammar (CG, \newcite{karlsson1995constraint})
% \citep[CG,][]{karlsson1995constraint} 
is a formalism used to disambiguate morphologically analysed text. 
A grammar consists of rules that target specific readings for selection or removal, based on contextual tests. For example, the following rule
\begin{itemize}
\item[] \texttt{REMOVE verb IF (-1 det) ;}
\end{itemize}
removes all verb readings from a word which is preceded by a determiner.
Given the following text,
\begin{itemize}
\item[] 
\begin{verbatim}
"<the>"
        "the" det def
"<wish>"
        "wish" noun sg
        "wish" verb pl
        "wish" verb inf
\end{verbatim}
\end{itemize}
the rule will match to the word \emph{wish}, and remove the readings
\texttt{"wish" verb pl} and \texttt{"wish" verb inf}.
Note that if the target word has only one remaining reading, then the
rule will not apply, even if the condition is met.

CGs are valuable resources for rule-based NLP, especially for lesser
resourced languages. They are robust and can be written without large
corpora---only a morphological analyser is needed. The formalism is
lightweight and language-independent, and resources can be shared
between related languages \cite{bick2006spanish,lene_trond_linda2010}.
Mature CGs contain some thousands of rules, but even small CGs are
shown to be effective \cite{lene_trond2011}.

@@

\begin{itemize}
\item Are all the rules distinct? (e.g. \texttt{Para} and \texttt{De} may be included in \texttt{Prep})
\item Can two or more rules be merged? (e.g. \texttt{SELECT Inf IF
    (-1C Prep OR Vai OR Vbmod)})
\item Can a messy rule be rewritten in a neater way without changing the meaning?
\item What is the best order for the rules?
\item Generate a sentence that triggers a given list of rules $R$ but not $R'$ 
\end{itemize}

@@

For the second use case, here are examples of conflicts that our tool will detect.
\begin{itemize}
\item If two equivalent rules $r$ and $r'$ occur in the grammar, the second occurrence will be disabled by the first
\item Rule $r$ selects something in a context, and $r'$ removes~it
\item A list of tules $R$ removes something from the context of a rule $r$, so $r$ can never
  apply
\item A rule $r$ has an internal conflict, such as non-existent
tag combination, or contradictory requirements for a context word
\end{itemize}
In the above examples, $R$ can be a single rule or a list of rules: for instance, if one rule removes a verb in
context $C$, and another in context $\neg C$, together these rules
remove a verb in all possible cases, disabling any future rule that
targets verbs.

@@

We express these constraints as a \emph{Boolean satisfiability problem} (SAT).
A SAT-problem consists of two components: a set of Boolean variables, and a set
of clauses on those variables. For instance, let the set of variables
be $\{a, b\}$ and the formulas $\{a \vee b, \neg{}a\}$. A program called a
\emph{SAT-solver} will try to find a solution, where all the variables
are replaced by a Boolean value. For this particular problem, the
unique solution is $\{a=False, b=True\}$, but it is also possible for a
SAT-problem to have no solution or multiple solutions.
In the case of CG analysis, the solution we build (called the \emph{model}) represents a \emph{sentence} that starts as potentially anything, and is being shaped by all the
rules along the way. This is how we can generate possible inputs, as
well as check if the grammar is internally consistent.

The paper is structured as follows. Section~\ref{sec:prev} relates our
work to previous work. 
Section~\ref{sec:implementation} discusses the implementation, and
Section~\ref{sec:eval} presents preliminary results. 
Section~\ref{sec:conclusion} discusses future work and concludes the paper.
